Rodent trapping studies to understand the prevalence of zoonotic disease in West Africa: A scoping review.

### 2021-08-27 reports/manuscript_v2.Rmd
updating map/analysis

### Rodent presence, absence, abundance
  * We mapped genus and species names to the species names used in the GLobal Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
  * We extacted information on the presence, absence and number of trapped individuals. For studies reporting on all trapped individuals (i.e. not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.

@@

  * We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted. In studies not reporting the year of trapping we used the year of publication as a proxy.
  * We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa. 
  * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites).
  * We summarised the habitat types of trap sites based on information reported in the study.
  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites was to SEDAC Global Population Density estimates (1970-2000 (10 year) and 2000-2020 (5 year)) and to land cover classifications (1975, 2000 and 2013) to explore the association of human population and rodent trapping effort.
  * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled. 
  * **Which statistical test to compare?** Ranked correlation to compare presence of habitat type against number of trap nights in habitat? Same analysis but weighted for human population?

@@

  * Rodent trapping took place across 1,307 study sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
  * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites. The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou). Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 1 - Panel B).
  * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
  * Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary). At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites. 
  * The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
  * Explore whether trap sites are more likely to have been conducted within high population regions or whether there is an association with trap location and habitat as obtained from remote sensing data (Need to do this. The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)

Figure 1 - Panel A - location of traps, Panel B - The distribution of population density at trap sites, Panel C - The distribution of habitat types at trap sites compared to all habitat types in West Africa

## Rodent presence, absence, abundance
  * 73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. 
  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-23%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
  * Of the 147 distinct identified species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112). Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
  * The most commonly trapped genera of rodents were Mastomys sp. (27,072, 38.5%), Rattus sp. (10,191, 14.5%), Mus sp. (8,624, 12.3%), Arvicanthis sp. (5,821, 8.3%) and Praomys sp. (5,409, 7.7%). At species level Mastomys natalensis (11,221, 17.9%), Rattus rattus (8,578, 13.7%), Mastomys erythroleucus (7,379, 11.8%), Mus musculus (6,245, 10%), Arvicanthis niloticus (5,497, 8.8%) and Mastomys huberti (4,699, 7.5%) were the most commonly trapped. 
  * Presence and absence maps for the six most commonly trapped species are compared to GBIF and IUCN data. (Need to do this)
  * The number of individuals trapped provides a measure of abundance, the interpretation of this is limited by incomplete reporting on the trapping effort (i.e. number of trap nights) within a study. The most abundant species at presence locations were … (Need to do this. I will explore the number of individuals trapped by species, I imagine it will show that if a species is detected at a site the most abundant ones will be the synanthropic species such as mus, rattus etc. while the more specialised species will be less abundant in areas where they are detected)

>>>>

### Rodent presence, absence, abundance
  * We mapped genus and species names to the species names used in the GLobal Biodiversity Information Facility (GBIF) taxonomy  [@gbif_the_global_biodiversity_information_facility_global_2021]
  * We extracted information on the presence, absence and number of trapped individuals. For studies reporting on all trapped individuals (i.e. not those only reporting on the presence of a specific species of interest), the pseudo-absence of a species reported as present elsewhere in the study was explicitly recorded as an absence at that trap location.

@@

  * We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted. In studies not reporting the year of trapping we used the year of publication as a proxy.
  * We used the GPS coordinates of single trap, trapping grid/line or study site and the number of trap nights to calculate trapping effort (trap night density) within level 2 administrative areas in West Africa. 
  * For studies not reporting the number of trap nights we imputed the number based on the median trap success rate from studies which reported the number of trap nights (matched to building or non-building based study sites). The median trap success rate for all rodents at a defined trap site was calculated for both trap sites including built environments and non-built environments. The number of rodents trapped at a trap site was then used in combination with these values of trap success to impute the number of trap nights for trap sites with no reported trap nights.
  * We summarised the habitat types of trap sites based on information reported in the study.
  * For the subset of studies investigating rodent zoonoses we compared the location of trapping sites was to SEDAC Global Population Density estimates (1970-2000 (10 year) and 2000-2020 (5 year)) and to land cover classifications (1975, 2000 and 2013) to explore the association of human population and rodent trapping effort.
  * We mapped the presence and absence of rodent species and compared this to the presence and absence reported by both GBIF and IUCN to give a measure of the extent of each species range in which they have been sampled. 
  * **Which statistical test to compare?** Ranked correlation to compare presence of habitat type against number of trap nights in habitat? Same analysis but weighted for human population?

@@

  * Rodent trapping took place across 1,331 trap sites with at least one trap site recorded from 14 West African countries. No rodent trapping studies were identified from Gambia or Togo.
  * Thirty-one (25%) studies reported trapping at a single study site, 46 (37%) studies trapped at between two and five study sites, the remaining 49 studies trapped at between six and 93 study sites. The areas with highest trapping activity included the capital cities of Niger (Niamey), Sierra Leone (Freetown), Senegal (Dakar), Mali (Bamako) and Ghana (Accra) and the largest cities of Ivory Coast (Abidjan) and Benin (Cotonou). Outside of these cities, Northern Senegal (Fatique, Thies, Saint-Louis and Kedougou), Southern Guinea (Kindia and Nzerejore), Edo and Osun States in Nigeria and Eastern Sierra Leone were the most intensively studied regions (see Figure 1 - Panel B).
  * No studies reported trap habitats with reference to a standardised habitat classification scheme. 
  * Extracted habitat types were grouped into 30 categories (see Supplementary Material 2 for the habitat dictionary). At least one habitat was recorded for 17,122 trap sites (95%), with two or more habitats for a single trap site recorded for 4,403 (24%) sites. Single trap sites could span multiple habitat types resulting in 22,202 distinct habitat and trap sites. 
  * The most commonly trapped sites were in or around buildings (29%), in areas described as the rodents “natural habitat” (24%), in agricultural areas (13%) (e.g. rice fields, palm plantations), forests (8%) and in the area surrounding buildings (4%).
  * Explore whether trap sites are more likely to have been conducted within high population regions or whether there is an association with trap location and habitat as obtained from remote sensing data (Need to do this. The remote sensing data will be matched to the trap dates there will be several years between the trapping and data this is compared to it seems impractical to try and get annual population estimates or land classification so this will be chosen pragmatically)

Figure 1 - Panel A - plot of studies, Panel B - location of traps, Panel C - The distribution of population density at trap sites, Panel D- The distribution of habitat types at trap sites compared to all habitat types in West Africa

Figure 1: **Panel A:** Each row represents one of the 126 included studies, green points designate the first year of data collection, blue points designate the end of data collection. For studies completed within one year the blue point overlies the green. Studies with a transparent grey point did not report the year in which trapping was conducted. The year of publication is shown by a red point. **Panel B:** Map of West Africa, countries where rodent trapping has occurred are mapped to level 2 administrative areas (where available). These regions are coloured by the total number of trap nights performed at trap sites within their boundaries. **Panel C:** Histogram of human population density at 5km^2^ resolution across West Africa and for areas where trapping has occurred. **Panel D:** Histogram of habitat types at 1km^2^ resolution across West Africa and for areas where trapping has occurred.

## Rodent presence, absence, abundance
  * 73,164 small mammals were trapped (592 were trapped outside of West African countries), 2,830 (4%) trapped individuals were identified to order level (Rodentia), 7,760 (11%) were identified to genus level, the remaining 62,574 (85%) were identified to species level. 
  * In studies reporting the number of trap nights the median trap success rate of traps placed in or around buildings was 13% (IQR 6-24%), this compares to a median trap success rate of 3% (IQR 1-9%) in other habitats.
  * Of the 147 distinct identified species trapped (see Supplementary Material 3) the majority were from the order Rodentia (112). Muridae (82) represented the largest family of rodents, followed by, Sciuridae (10), Gliridae and Nesomyidae (both 6), Ctenodactylidae and Anomaluridae (2) and Dipodidae, Hystricidae and Thryonomidae (all 1). The remaining 34 species came from the orders of the Soricomorpha (30), Erinaceomorpha (2) and Afrosoricida (1).
  * The most commonly trapped genera of rodents were Mastomys sp. (27,072, 38.5%), Rattus sp. (10,191, 14.5%), Mus sp. (8,624, 12.3%), Arvicanthis sp. (5,821, 8.3%) and Praomys sp. (5,409, 7.7%). At species level Mastomys natalensis (11,221, 17.9%), Rattus rattus (8,578, 13.7%), Mastomys erythroleucus (7,379, 11.8%), Mus musculus (6,245, 10%), Arvicanthis niloticus (5,497, 8.8%) and Mastomys huberti (4,699, 7.5%) were the most commonly trapped. 
  * Presence and absence maps for the six most commonly trapped species are compared to GBIF and IUCN data. (Need to do this)
  * The number of individuals trapped provides a measure of abundance, the interpretation of this is limited by incomplete reporting on the trapping effort (i.e. number of trap nights) within a study. The most abundant species at presence locations were … (Need to do this. I will explore the number of individuals trapped by species, I imagine it will show that if a species is detected at a site the most abundant ones will be the synanthropic species such as mus, rattus etc. while the more specialised species will be less abundant in areas where they are detected)
